Stack Safety is a Security Property

### 2021-04-12 tex/main.tex
Addressed notes in abstract, introduction, some in section 2

We propose a formal criterion for stack safety, phrased in terms of
language-based security: a combination of an integrity property (``each
caller's stack frame is invariant by the callee'') and a confidentiality
property (``the callee's behavior is insensitive to differences in the
caller's state''). These are defined in terms of a control flow model
in which callees are meaningfully distinct from their callers due to being
bracketed by calls and returns. We further apply a straightforward notion of
``well-bracketed control flow'' that ensures that callers' control flow
can continue predictably after the return.

  \bcp{I am still not clear
  how well-bracketedness fits into the story.  (For me, it is the bracketing
  that {\em defines} where calls and returns happen, so there is something
  tautological about saying that `` well-bracketedness condition ensures
  that callers and callees remain meaningfully distinct''!)}
  \sna{Attempted to resolve this, it may need some work.}

Our criterion is inspired by hardware tagging mechanisms for stack
protection, in particular the stack-safety ``micro-policies''
of~\citet{DBLP:conf/sp/RoesslerD18}, but its core reflects general
principles of stack safety\bcp{I wish we did not have to be so apologetic
  about being tied to micro-policies (and it is still not clear to me that
  we have to be... do we have a clear statement someplace of why our
  definition would not apply to CFI, for example? This seems like a critical
  discussion...)}. Finally, we show that micro-policies do indeed enforce
our properties using property-based randomized testing. An eager micropolicy
that catches violations immediately enforces our stronger property, and we
define a weaker, more extensional form of the property that is enforced
by a ``lazy'' micropolicy -- one that, for performance reasons, may permit
local violations as long as they do not influence future behavior.

@@

The call stack is a perennial target for low-level attacks
based on memory-safety vulnerabilities, leading to consequences
from leakage or corruption of private stack data to
control-flow hijacking. To detect or prevent such attacks, a profusion of
software and hardware protections have been proposed,
%
including stack canaries~\citep{Cowan+98},
bounds checking~\citep{NagarakatteZMZ09,NagarakatteZMZ10,DeviettiBMZ08},
split stacks~\citep{Kuznetsov+14},
shadow stacks~\citep{Dang+15,Shanbhogue+19},
capabilities~\citep{Woodruff+14,Chisnall+15,Skorstengaard+19,Skorstengaard+19b,Tsampas+19,Georges+21},
and hardware tagging~\citep{DBLP:conf/sp/RoesslerD18}. \ifaftersubmission\apt{Mostly from
  nick; there could be more}\bcp{Yes, going back to MIT days---we should
  include several more of these, if only to give readers the impression that
this is a well-studied mechanism (so formalizing its protections is
useful).}
\fi
%
The protections offered by such mechanisms are generally described in terms
of a set of concrete examples of attacks that they can prevent, leaving a
more general characterization to the reader's intuition.
Moreover, the details of these mechanisms can be quite intricate
and often involve trading off protection and efficiency.
Thus, it would be useful to have a precise, generic, and formal
specification for stack
safety, both as a basis for comparing the security claims of different
enforcement techniques and for validating more rigorously that these claims
are met by particular implementations.

So how should we formally define stack safety? Definitions in the literature are commonly
informal and proceed by enumerating the specific bad things that a given enforcement mechanism
prevents, such as corruption of return addresses, buffer overflows, or
use of uninitialized variables.\bcp{Most of that sentence should be merged
  into the similar sentence above.  Doesn't belong here.}
%\bcp{Does use of unitialized variables lead to stack safety attacks?\apt{sure, via buffer overflow}}
%with the notion usually being
%defined negatively: through examples of stack-based exploits
%\rb{usually though not always; as noted below there are other
%  approaches}\bcp{Yes, we have to be very careful about this}.
%\leo{Examples at a high level}: such as -- does not overwrite return addresses, or ...
The most abstract characterization of which we are aware is in
recent work by \citet{Skorstengaard+19}, which defines stack safety as the conjunction of two
properties:
{\em local state encapsulation (LSE)} and {\em well-bracketed control flow (WBCF)}.
Informally, LSE says that the contents of
a caller's stack frame are not read and written while its callees (and their callees, etc.) are
executing, and WBCF says that callees always
return to the point just after their call (or don't return at all).
\citet{Skorstengaard+19} formalize these properties by defining an idealized machine
in which they visibly hold by construction; an enforcement mechanism (in their case, hardware
capabilities)
can then be shown to provide stack safety by giving a fully abstract embedding function
from the idealized machine to an ordinary machine equipped with the enforcement.
%%
%% While this work represents an important step in the right direction, we believe
%% there is still room to craft an operational characterization
%% of stack safety. \apt{Perhaps our rhetoric should be more micro-policy-specific here?}
%% In other words, we seek formal criteria that can be applied
%% to a mechanism for protecting stack-based low-level programs to judge
%% whether or not it actually ``guarantees stack safety.''
%% \rb{The remainder of this paragraph may be diving too quickly, would a reader be
%% confused by the references to these types of properties, yet to be introduced?}
%% We begin with a strong
%% local state encapsulation property that closely models the behavior of eager
%% micro-policies from \citet{DBLP:conf/sp/RoesslerD18}. This ``eager'' property
%% is too strong to characterize stack safety; reasonable enforcement mechanisms
%% exist that violate it. One such mechanism is a ``lazy'' tag policy, similar to
%% the one in \citet{DBLP:conf/sp/RoesslerD18}. We weaken local state encapsulation
%% to support this technique and argue that the new property, defined relative to an
%% arbitrary notion of observation. This property is still strong enough to prevent
%% the bad behaviors we associate with unsafe stack behavior, but weak enough that
%% other reasonable enforcement mechanisms can enforce it.

\bcp{If we end up adopting my recent proposal for defining subtraces and
  separating WBCF, I think we can be a bit stronger here: We define stack
  safety in IFC terminology based on the same intuitions as their LSE, but
  we consider (and define) WBCF to be a separate thing---desirable, but
  distinct from safety.  (If we make this argument, we'll need to show an
  example of a program that is stack safe but doesn't have WBCF.)}
In this work, we take the (informal) definitions of WBCF and LSE as a starting
point, but choose a different approach to formalizing them, based on the
technical framework of language-based security~\citep{??}. We define
trace-based properties of \emph{stepwise stack integrity} and
\emph{lockstep stack confidentiality}.
In applying these security concepts, we treat each caller-callee pair as
mutually distrustful, with the caller needing protection from the callee. \apt{Need more explanation of
  ``nesting'' and our novelty factor here? Also about domains??}
This differs from common instances in language-based security in that the
confidentiality property applies to many nested subtraces within a single execution.
\sna{Thinking about how to express domains, or whether to. I'm not sure it's
  an interesting detail at this level. Also not sure how strongly to talk about this novelty.}
{\em Integrity} says that a caller's local data is protected from modification by its callee,
and its {\em stepwise} interpretation formalizes this very naturally as a caller's local data
never being modified during a call. {\em Confidentiality} says that a caller's data is protected
from being leaked into its callee, and also that the callee cannot see uninitialized data.
As a reasoning principle, confidentiality guarantees that the callee's
behavior is determined only by a predictable interface that does not include the caller's
local state or the uninitialized portion of the stack. This is formalized as a
{\em noninterference}-style property that, rather than comparing pairs of whole program traces,
compares a pair of subtraces for each call, nested within other calls. The term
{\em lockstep} describes the behavior of these pairs: they step together, and any
deviation is recognized as a violation of confidentiality.
Together, these properties capture the rough intuition of LSE, but are not identical --
confidentiality contains temporal protections that LSE does not address.

@@

In summary, we offer the following contributions: \bcp{Section numbers are
  wonky---why just 7.1, where is 8, etc.}
\begin{itemize}
\item
  We adapt formulations of integrity and confidentiality from
  language-based security
  to formalize a parameterized {\em stepwise stack-safety property} and argue that
  it captures the intuitive concepts of local stack encapsulation, well-bracketed control
  flow, and temporal stack safety~(\cref{sec:lse-and-wbcf}).
\item
  We use a property-based randomized testing framework to support the claim
  that this stepwise property is actually enforced by the Depth Isolation policy
  of~\citet{DBLP:conf/sp/RoesslerD18}~(\cref{sec:conservative}).
\item
  We formalize a weaker but more extensional \emph{observational} variant of the
  stack safety property and argue that it accurately characterizes
  the Lazy Tagging and Clearing policy of~\citet{DBLP:conf/sp/RoesslerD18}
  enhanced with unique tags on each activation~(\cref{sec:lazy}).
\item
  We show how our models can be instantiated and extended to handle passing
  arguments on the stack~(\cref{sec:??}) and simple
  coroutines~(\cref{sec:??}).\bcp{``instantiated and extended'' is awkward.
  Maybe ``We show how our models can be instantiated to handle passing
  arguments on the stack~(\cref{sec:??}) and extended with simple
  coroutines~(\cref{sec:??}).''  If that's what we mean.}
%  This policy does not enforce our strict
%  local state encapsulation property,
%  but a modified version of it using unique per-activation tags enforces
%  an (also provably) weaker lazy variant of the property, which still prohibits
%  the same intuitively stack unsafe behaviors.
%
% \apt{Awkward, but the truth I fear.} \bcp{I agree the phrasing is awkward,
%   but the point we can make is still strong: To make the property easier
%   to enforce, we can slightly weaken its guarantees; the weaker version is
%   a bit more complicated to state but still retains the pragmatic
%%   protections of the original.}\apt{To clarify: Nick's LT\&C policy uses
%   depth-based tags, which do not enforce even our lazy policy. So we need to
%   introduce a version of LT\&C that uses unique activation tags.}

  %\item As evidence that the property is not too strong\apt{But we add the lazy version because the strict version \emph{is} too strong, in some sense.}, we show that it is
%  realizible by
%  a previously proposed micro-policy enforcement
%  mechanism~(\Cref{sec:micro-policy}). \leo{And it says yes on a
%    substantial body of code}. We also present a variation of the
%  property + policy that is lazy, easier/more efficient to implement,
%  and provably weaker (\Cref{sec:lazy}).
%\item A provably stronger (inductive? step-by-step/stepwise) variant
%  of this property tuned for testing/verification.  We test the whole
%  thing with a cool framework. \bcp{Which section?}
\end{itemize}
\Cref{sec:relwork} discusses related work, and
\cref{sec:future} sketches directions for future work.

We begin with a description of our assumptions and threat model (\cref{sec:threat}),
then show how our properties apply to examples of stack unsafety 
(\cref{sec:running-example}). Then, after some technical preliminaries
(\cref{sec:prelim}), we formalize the properties (\cref{sec:eager}, \cref{sec:lazy})
and describe testing (\cref{sec:testing}).

@@

This division of policy state from machine state is helpful when reasoning about
enforcement mechanisms that keep a significant amount of data separate from the
primary execution data to inform security. Obviously tag-based systems do this,
with tags and extra policy state separated from and untouchable by application code.
Another significant example is the shadow stack, where a second stack
mirrors the real one, and a return may only occur if both stacks agree on the
target address. Typically the shadow stack is isolated using a separate protection
mechanism.

@@

Intuititively, stack safety means that a caller's stack frame is protected
from a confused or malicious callee's attempts to read or write its private data.
The attacker model therefore treats every caller as a defender and its callee as
an attacker, meaning that with nested calls, a single function may play both roles
at varying points.

@@

%\leo{Anything else to add here?}
%\leo{Maybe a good place to discuss calling conventions and that they
%  are essentially enforced by the policy or sth?}
%\rb{Intuition that the callee can do anything in principle, but the machine will
%halt if it does not follow the calling convention, etc.?}

@@

Since we rely heavily on code annotations, we must ask: what if the annotations
are wrong? First, suppose a call or return occurs that is not annotated.
In this case, the answer is simple: as far as the properties are concerned, there has
been no call or return, only an instruction sequence resembling such an operation, and
therefore no associated change in permissions. This is particularly interesting in the
case of a false return, where control may appear to have returned to the caller, but
the enforcement mechanism must still protect the caller's memory.

>>>>

We propose a formal criterion for stack safety, phrased in terms of
language-based security: a combination of an integrity property (``each
caller's stack frame is invariant by the callee'') and a confidentiality
property (``the callee's behavior is insensitive to differences in the
caller's state''). These properties are sensible even in the presence of
abnormal control flow. We further apply a straightforward notion of
``well-bracketed control flow'' that restricts control flow to the
typical pattern matched calls and returns.

We show that our properties describe the behavior the stack-safety
``micro-policies'' of~\citet{DBLP:conf/sp/RoesslerD18}
using property-based randomized testing. An eager micropolicy
that catches violations immediately enforces our primary ``stepwise''
properties, and we define weaker, more extensional ``observational''
properties that are enforced even by a ``lazy'' micropolicy -- one
that, for performance reasons, may catch violations after they occur
but before they become visible.

We argue that our properties are suited to a range of other enforcement
mechanisms. The strict stepwise properties can describe mechanisms
that catch errors immediately, as is typical. While the observational
properties are closely linked to lazy micropolicies they are weak enough
to be a lower bound that captures stack safety for any reasonable enforcement.

@@

The call stack is a perennial target for low-level attacks
based on memory-safety vulnerabilities, leading to consequences
from leakage or corruption of private stack data to
control-flow hijacking. To detect or prevent such attacks, a profusion of
software and hardware protections have been proposed,
%
including stack canaries~\citep{Cowan+98},
bounds checking~\citep{NagarakatteZMZ09,NagarakatteZMZ10,DeviettiBMZ08},
split stacks~\citep{Kuznetsov+14},
shadow stacks~\citep{Dang+15,Shanbhogue+19},
capabilities~\citep{Woodruff+14,Chisnall+15,Skorstengaard+19,Skorstengaard+19b,Tsampas+19,Georges+21},
and hardware tagging~\citep{DBLP:conf/sp/RoesslerD18}. \ifaftersubmission\apt{Mostly from
  nick; there could be more}\bcp{Yes, going back to MIT days---we should
  include several more of these, if only to give readers the impression that
this is a well-studied mechanism (so formalizing its protections is
useful).}
\fi
%
The protections offered by such mechanisms are generally described in terms
of a set of concrete examples of attacks that they can prevent, such as corruption
of return addresses, buffer overflows, or use of uninitialized variables. This leaves a
more general characterization to the reader's intuition.
Moreover, the details of these mechanisms can be quite intricate
and often involve trading off protection and efficiency.
Thus, it would be useful to have a precise, generic, and formal
specification for stack
safety, both as a basis for comparing the security claims of different
enforcement techniques and for validating more rigorously that these claims
are met by particular implementations.

%\bcp{Does use of unitialized variables lead to stack safety attacks?\apt{sure, via buffer overflow}}
%with the notion usually being
%defined negatively: through examples of stack-based exploits
%\rb{usually though not always; as noted below there are other
%  approaches}\bcp{Yes, we have to be very careful about this}.
%\leo{Examples at a high level}: such as -- does not overwrite return addresses, or ...
The most abstract characterization of which we are aware is in
recent work by \citet{Skorstengaard+19}, which defines stack safety as the conjunction of two
properties:
{\em local state encapsulation (LSE)} and {\em well-bracketed control flow (WBCF)}.
Informally, LSE says that the contents of
a caller's stack frame are not read and written while its callees (and their callees, etc.) are
executing, and WBCF says that callees always
return to the point just after their call (or don't return at all).
\citet{Skorstengaard+19} formalize these properties by defining an idealized machine
in which they visibly hold by construction; an enforcement mechanism (in their case, hardware
capabilities)
can then be shown to provide stack safety by giving a fully abstract embedding function
from the idealized machine to an ordinary machine equipped with the enforcement.
%%
%% While this work represents an important step in the right direction, we believe
%% there is still room to craft an operational characterization
%% of stack safety. \apt{Perhaps our rhetoric should be more micro-policy-specific here?}
%% In other words, we seek formal criteria that can be applied
%% to a mechanism for protecting stack-based low-level programs to judge
%% whether or not it actually ``guarantees stack safety.''
%% \rb{The remainder of this paragraph may be diving too quickly, would a reader be
%% confused by the references to these types of properties, yet to be introduced?}
%% We begin with a strong
%% local state encapsulation property that closely models the behavior of eager
%% micro-policies from \citet{DBLP:conf/sp/RoesslerD18}. This ``eager'' property
%% is too strong to characterize stack safety; reasonable enforcement mechanisms
%% exist that violate it. One such mechanism is a ``lazy'' tag policy, similar to
%% the one in \citet{DBLP:conf/sp/RoesslerD18}. We weaken local state encapsulation
%% to support this technique and argue that the new property, defined relative to an
%% arbitrary notion of observation. This property is still strong enough to prevent
%% the bad behaviors we associate with unsafe stack behavior, but weak enough that
%% other reasonable enforcement mechanisms can enforce it.

In this work, we take the (informal) definitions of WBCF and LSE as a starting
point, but choose a different approach to formalizing them, based on the
technical framework of language-based security~\citep{??}. We define
trace-based properties of \emph{stepwise stack integrity} and
\emph{lockstep stack confidentiality}.
In applying these security concepts, we treat each caller as needing protection
from its callee, even as that callee may need protection from its callee in turn.
This differs from common instances in language-based security in that the
confidentiality property applies to many nested subtraces within a single execution.
\sna{How novel is this nesting?}
{\em Integrity} says that a caller's local data is protected from modification by its callee,
and its {\em stepwise} interpretation formalizes this very naturally as a caller's local data
never being modified during a call. {\em Confidentiality} says that a caller's data is protected
from being leaked into its callee, and also that the callee cannot see uninitialized data.
As a reasoning principle, confidentiality guarantees that the callee's
behavior is determined only by a predictable interface that does not include the caller's
local state or the uninitialized portion of the stack. This is formalized as a
{\em noninterference}-style property that, rather than comparing pairs of whole program traces,
compares a pair of subtraces for each call, nested within other calls. The term
{\em lockstep} describes the behavior of these pairs: they step together, and any
deviation is recognized as a violation of confidentiality.
Together, these properties capture the rough intuition of LSE, but are not identical --
confidentiality contains temporal protections that LSE does not address.

@@

In summary, we offer the following contributions: \bcp{Section numbers are
  wonky---why just 7.1, where is 8, etc.}
\begin{itemize}
\item
  We adapt formulations of integrity and confidentiality from
  language-based security
  to formalize a parameterized {\em stepwise stack-safety property} and argue that
  it captures the intuitive concepts of local stack encapsulation, well-bracketed control
  flow, and temporal stack safety~(\cref{sec:lse-and-wbcf}).
\item
  We use a property-based randomized testing framework to support the claim
  that this stepwise property is actually enforced by the Depth Isolation policy
  of~\citet{DBLP:conf/sp/RoesslerD18}~(\cref{sec:conservative}).
\item
  We formalize a weaker but more extensional \emph{observational} variant of the
  stack safety property and argue that it accurately characterizes
  the Lazy Tagging and Clearing policy of~\citet{DBLP:conf/sp/RoesslerD18}
  enhanced with unique tags on each activation~(\cref{sec:lazy}).
\item
  We extend our model so that it can be instantiated in systems that pass arguments
  on the stack~(\cref{sec:??}) and in those with simple coroutines~(\cref{sec:??}).
\end{itemize}

We begin with a description of our assumptions and threat model (\cref{sec:threat}),
then show how our properties apply to examples of stack unsafety 
(\cref{sec:running-example}). Then, after some technical preliminaries
(\cref{sec:prelim}), we formalize the properties (\cref{sec:eager}, \cref{sec:lazy})
and describe testing (\cref{sec:testing}).
\Cref{sec:relwork} discusses related work, and
\cref{sec:future} sketches directions for future work.

@@

\sna{This is awkward.}
This division of policy state from machine state is helpful when reasoning about
enforcement mechanisms that keep a significant amount of data separate from the
primary execution data to inform security. Obviously tag-based systems do this,
with tags and extra policy state separated from and untouchable by application code.
Another significant example is the shadow stack, where a second stack
mirrors the real one, and a return may only occur if both stacks agree on the
target address. Typically the shadow stack is isolated using a separate protection
mechanism.

@@

Intuitively, stack safety means that a caller's stack frame is protected
from a confused or malicious callee's attempts to read or write its private data.
The attacker model therefore treats every caller as a defender and its callee as
an attacker, meaning that with nested calls, a single function may play both roles
at varying points.

@@

@@

\sna{TODO: Need to rewrite this for the new model of WBCF.}
Since we rely heavily on code annotations, we must ask: what if the annotations
are wrong? First, suppose a call or return occurs that is not annotated.
In this case, the answer is simple: as far as the properties are concerned, there has
been no call or return, only an instruction sequence resembling such an operation, and
therefore no associated change in permissions. This is particularly interesting in the
case of a false return, where control may appear to have returned to the caller, but
the enforcement mechanism must still protect the caller's memory.
